---
title: "FishBase data exploration"
output:
  html_document:
    theme: paper
---

```{r setup, include=FALSE}
## script to explore fishbase data 
## developed by Nikki

library(tidyverse)
library(data.table)
library(ggpubr)
library(kableExtra)
library(patchwork)

# set ggplot themes
theme_set(theme_pubclean())

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE) 

# make palette
colors <- c("All Mammals" = "slategray",
            "Canadian Wild Species" = "navyblue",
            "C-LPI" = "firebrick")

## read in data:
clpi_fish <- fread('../data-clean/clpi_fishbase_merge.csv')
all_traits <- fread('../data-clean/fishbase_all-spp.csv')
can_traits <- fread('../data-clean/fishbase_canadian-spp.csv')

## subset to just traits:
traits <- clpi_fish %>%
  select(Binomial, Genus, Species, c(89:118)) %>%
  unique()

## write function silently:
summarise_traits <- function(spp_traits) {
  ## look at all the values of our traits
  trait_summary <- spp_traits %>%
    select("LongevityWild","LongevityCaptive", "Length", 
           "LTypeMaxM", 
           "LengthFemale", 
           "LTypeMaxF",
           "CommonLength", 
           "LTypeComM", 
           "CommonLengthF",
           "LTypeComF",
           "Weight", "WeightFemale", 
           "Median_T", "lcl_T", "ucl_T",
           "Herbivory2", "DietTroph", "DietSeTroph", 
           "FoodTroph", "FoodSeTroph",
           "Troph", "seTroph") %>%
    summarise_all(~ sum(is.na(.))) %>%
    gather(key = "Trait", value = 'Number of species missing trait') %>%
    mutate('Variable_type' = c(rep("continuous", 3), 
                               "NA", 
                               "continuous",
                               "NA",
                               "continuous",
                               "NA", 
                               "continuous",
                               "NA",
                               rep("continuous", 5),
                               'NA',
                               rep("continuous", 6)
    )) %>%
    mutate(Units = c(rep("years", 2), 
                     "cm", "NA", 
                     "cm",
                     "NA",
                     "cm",
                     "NA", 
                     "cm",
                     "NA", 
                     rep("grams", 2),
                     rep('years', 3),
                     "NA",
                     rep('trophic levels', 6)))
  
  pasted = data.frame(val_cat = sapply(spp_traits, function(x) paste(unique(x),
                                                                     collapse = ', '))) %>%
    mutate(var = rownames(.)) %>%
    filter(var %in% trait_summary$Trait)
  
  minmax = data.frame(val_cont = sapply(spp_traits, 
                                        function(x) paste(round(min(as.numeric(as.character(x)),
                                                                    na.rm=T),2),
                                                          round(max(as.numeric(as.character(x)),
                                                                    na.rm=T),2),
                                                          sep = ' - '))) %>%
    mutate(var = rownames(.)) %>%
    filter(var %in% trait_summary$Trait)
  
  vals = left_join(pasted, minmax) 
  
  trait_summary = trait_summary %>%
    left_join(., vals, by = c("Trait" = 'var')) %>%
    mutate('Values' = ifelse(Variable_type == 'continuous', as.character(val_cont),
                             as.character(val_cat))) %>%
    select(-val_cont, -val_cat) %>%
    select(Trait, `Variable_type`, Units, Values, everything())
  
  return(trait_summary)
}


```

Trait scripts for fish used in this document are created `01_querying-fishbase.R`

## Trait data completeness & cleanliness

How complete are the fish traits we want for our species?
\n
```{r complete, echo = FALSE}
## how complete are traits?
nas <- data.frame(trait = names(colSums(is.na(traits))), nas = colSums(is.na(traits))) %>%
  mutate(prop_missing = as.numeric(as.character(nas))/368) 

nas %>%
  filter(trait %in% c("LongevityWild", "Length", "CommonLength", "DietTroph", "FoodTroph", "Troph",
                      "Median_T","Weight")) %>%
  ggplot(., aes(x = trait, y = prop_missing*100, fill = trait)) + geom_col() +
  labs(x = "Trait", y = '% of CLPI fish species missing trait value') +
  scale_x_discrete(labels = c("Common length (cm)", "Max length (cm)", "Longevity in wild (years)", 
                              "Median generation time (years)", 
                              "Trophic position (actual diet composition studies)",
                              "Trophic position (based on known food items)", 
                              "Trophic position (model estimate)", "Weight")) +
  guides(fill = F) +
  coord_flip() +
  theme_minimal()
```

\n\n
How clean are the traits?
```{r clean, echo = FALSE}
kable(summarise_traits(traits)) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")
```


\n
\n
### Distributions of measured traits:

```{r,fig.show='hold', message=FALSE, warning = FALSE, fig.height = 4, fig.width=9}
A <- ggplot(traits) +
  geom_histogram(aes(x = log(Weight)),
                 bins = 15, col = "white",
                 fill = "navyblue", alpha = .6) +
  labs(x = "log weight (g)")
B <- ggplot(traits) +
  geom_histogram(aes(x = log(Length)),
                 bins = 15, col = "white",
                 fill = "firebrick", alpha = .6) +
  labs(x = "Max length (cm)")
C <- ggplot(traits) +
  geom_histogram(aes(x = LongevityWild),
                 bins = 15, col = "white",
                 fill = "darkgreen", alpha = .6) +
  labs(x = "Max longevity in wild (years)")
A + B + C
```



### Distributions of model-estimated traits:

```{r,fig.show='hold', message=FALSE, warning = FALSE, fig.height = 4, fig.width=9}
A <- ggplot(traits) +
  geom_histogram(aes(x = Troph),
                 bins = 15, col = "white",
                 fill = "navyblue", alpha = .6) +
  labs(x = "Model estimated trophic position")
B <- ggplot(traits) +
  geom_histogram(aes(x = Median_T),
                 binwidth = 5, col = "white",
                 fill = "firebrick", alpha = .6) +
  labs(x = "Generation time (years)")
A + B 
```


\n
### How different are the measures of trophic level from one another?

```{r,fig.show='hold', message=FALSE, warning = FALSE, fig.height = 4, fig.width=9}
ggplot() +
  geom_density(data = traits,
               aes(x = DietTroph, fill = "Based on diet composition studies"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = traits,
               aes(x = FoodTroph, fill = "Based on known food items"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = traits,
               aes(x = Troph, fill = "Model estimated"),
               lwd = 0.2, alpha = .5) +
  labs(fill = "", x = "Trophic level")
```


\n
### How different are all of the different body length trait measurements?

```{r,fig.show='hold', message=FALSE, warning = FALSE, fig.height = 4, fig.width=9}
ggplot() +
  geom_density(data = traits,
               aes(x = log(Length), fill = "Max length"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = traits,
               aes(x = log(LengthFemale), fill = "Max length (female)"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = traits,
               aes(x = log(CommonLength), fill = "Common length"),
               lwd = 0.2, alpha = .5) +
   geom_density(data = traits,
               aes(x = log(CommonLengthF), fill = "Common length (female)"),
               lwd = 0.2, alpha = .5) +
  labs(fill = "", x = "log (length) (cm)")
```

\n
### Within the trait 'maximum length', how much does the measurement type matter? (ex. tail length vs. standard length vs. etc.)
```{r,fig.show='hold', message=FALSE, warning = FALSE, fig.height = 4, fig.width=9}
ggplot() +
  geom_density(data = traits,
               aes(x = log(Length), fill = LTypeMaxM),
               lwd = 0.2, alpha = .5) +
  labs(fill = "", x = "log (length) (cm)")
```


## Looking at some relationships between traits

Is there a correlation between trophic position and maximum body length?
```{r trophic, echo = FALSE}
traits %>%
  mutate(upper = Troph + seTroph, lower = Troph - seTroph) %>%
  ggplot(., aes(y = Troph, x = log10(Length), col = LongevityWild)) + geom_point() + 
  labs(y = 'Trophic position (model estimated)', x = 'Log (maximum body length) (cm)', 
       col = "Max. longevity (years)") + 
  geom_errorbar(aes(ymin = lower, ymax = upper))
```


Is there a correlation between maximum longevity and generation time?
```{r gentime, echo = FALSE}
traits %>%
  ggplot(., aes(y = Median_T, x = LongevityWild)) + geom_point() + 
  labs(y = 'Median generation time (years)', x = "Maximum longevity in wild (years)") +
  geom_errorbar(aes(ymin = lcl_T, ymax = ucl_T))
```




### Comparing trait distributions

These are the distributions of each trait for fish in the Canadian LPI dataset (**C-LPI**), compared with the distribution of these traits for all Canadian fish in the **Wild Species List**, and with all fish in the full trait dataset (**FishBase**).

```{r}
data.frame(
  "Scale" = c("Global", "Canada", "Canada (LPI)"),
  "Species list" = c("FishBase", "Wild Species Report", "Canadian-LPI"),
  "Number of fish species" = c(length(unique(all_traits$Binomial)), 
                                 length(unique(can_traits$Binomial)), 
                                 length(unique(traits$Binomial)))
  ) %>% 
  kableExtra::kable(col.names = c("Scale", "Species list source", "Number of fish species")) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")
```

***Note***: there are 18 fish species in the C-LPI dataset that **are not** in the Wild Species List!

```{r, echo = FALSE}
ggplot() +
  geom_density(data = all_traits,
               aes(x = log(LongevityWild), fill = "All fish"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = can_traits,
               aes(x = log(LongevityWild), fill = "Canadian Wild Species"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = traits,
               aes(x = log(LongevityWild), fill = "C-LPI"),
               lwd = 0.2, alpha = .5) +
   scale_fill_manual(values = colors) +
  labs(title = "Log (maximum longevity) (years)", x = "Log (maximum longevity)  (years)",
       y = "Density",
       fill = "Dataset")

ggplot() +
  geom_density(data = all_traits,
               aes(x = log(Length), fill = "All fish"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = can_traits,
               aes(x = log(Length), fill = "Canadian Wild Species"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = traits,
               aes(x = log(Length), fill = "C-LPI"),
               lwd = 0.2, alpha = .5) +
   scale_fill_manual(values = colors) +
  labs(title = "Log (maximum length) (cm)", x = "Log (maximum length) (cm)", y = "Density",
       fill = "Dataset")

ggplot() +
  geom_density(data = all_traits,
               aes(x = log(Weight), fill = "All fish"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = can_traits,
               aes(x = log(Weight), fill = "Canadian Wild Species"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = traits,
               aes(x = log(Weight), fill = "C-LPI"),
               lwd = 0.2, alpha = .5) +
   scale_fill_manual(values = colors) +
  labs(title = "Log (weight) (g)", x = "Log (weight) (g)", y = "Density",
       fill = "Dataset")

ggplot() +
  geom_density(data = all_traits,
               aes(x = Troph, fill = "All fish"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = can_traits,
               aes(x = Troph, fill = "Canadian Wild Species"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = traits,
               aes(x = Troph, fill = "C-LPI"),
               lwd = 0.2, alpha = .5) +
   scale_fill_manual(values = colors) +
  labs(title = "Model-estimated trophic position",
       x = "Model-estimated trophic position", y = "Density",
       fill = "Dataset")



```