---
title: "FishBase Data Extraction"
author: "Sandra Emry"
date: "31/08/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pkgs <- c("tidyverse", "rfishbase", "skimr")
lapply(pkgs, library, character.only = TRUE)
rm(pkgs)

wildsp <- readr::read_csv("/Users/sandraemry/Documents/LPI_working_group/trait_based/trait_data_hunting/data-raw/WildSpecies2015Data.csv", col_names = FALSE)

names(wildsp) <- as.character(wildsp[1, ])
canadian_fish_sp <- wildsp[2:nrow(wildsp),6]
canadian_fish_sp <- unique(canadian_fish_sp)
canadian_fish_sp <- canadian_fish_sp %>% 
  separate(Binomial, into = c("genus", "species"), sep = "_") %>% 
  unite("Binomial", c("genus", "species"), sep = " ", remove = TRUE)

#vector of all Canadian fish species 
canadian_fish_sp <- pull(canadian_fish_sp, Binomial)


```


Using rFishbase to pull trophic, length and age data for all fish species that are listed in the WildSpecies2015 Report that have data in fishbase. 

## Trophic Trait Data 
Below is some summary data for information extracted from the ecology() functin 
```{r}
#Pulling trophic level traits for all Canadian fish 
all_canadian_fish_trophic_data <- ecology(species_list = canadian_fish_sp,
                        fields = c("Species", "Herbivory2", "DietTroph", "DietSeTroph", 
                                   "FoodTroph", "FoodSeTroph", "DietRemark"))

## filter to species that have trophic information in at least one column 
all_canadian_fish_trophic_data <- all_canadian_fish_trophic_data %>% 
  filter_at(vars(Herbivory2, DietTroph, FoodTroph), any_vars(!is.na(.)))

skim(all_canadian_fish_trophic_data)

hist(all_canadian_fish_trophic_data$DietTroph)

```
## Length Data 
There is length information in multiple places in fishbase: length_weight(), popchar()
```{r}
all_canadian_fish_size_LW <- length_weight(canadian_fish_sp) %>% 
  select(Species, LengthMin, LengthMax, Type, Sex) %>% 
  filter_at(vars(LengthMin, LengthMax), any_vars(!is.na(.))) %>% ## filter to species that have size info in at least one column 
  rename(LengthMin_LW = LengthMin, LengthMax_LW = LengthMax, Type_LW = Type, Sex_LW = Sex)

skim(all_canadian_fish_size_LW)

all_canadian_fish_popchar <- popchar(canadian_fish_sp) %>% 
  select(Species, PopCharRefNo, Sex, Lmax, Type, tmax) %>% 
  rename(Lmax_popchar = Lmax, Type_popchar = Type, Sex_popchar = Sex, Life_Span_popchar = tmax)

skim(all_canadian_fish_popchar)
  
hist(as.numeric(all_canadian_fish_popchar$Lmax_popchar)) 
hist(all_canadian_fish_size_LW$LengthMax_LW)


```

## Life Span Data 
```{r}
all_canadian_fish_age <- estimate(canadian_fish_sp) %>% 
  select(Species, AgeMin, AgeMax) %>% 
  filter(!is.na(AgeMax))

all_canadian_fish_age[which(all_canadian_fish_age$AgeMax == max(all_canadian_fish_age$AgeMax)), ]
hist(all_canadian_fish_age$AgeMax)
hist(all_canadian_fish_popchar$Life_Span_popchar)

```

## Canadian LPI Fish Species
Now I'm going to pull the same types of data from fishbase but only for fish species that are also in the Canadian LPI Database
```{r}
# load Canadian LPI data
clpi_data <- read_csv("/Users/sandraemry/Documents/LPI_working_group/trait_based/trait_data_hunting/data-raw/CIEE_LPI_dataset.csv")

#chnage subclass to class
clpi_data$Class <- gsub("Actinopteri", "Fish",clpi_data$Class)
clpi_data$Class <- gsub("Coelacanthi", "Fish",clpi_data$Class)
clpi_data$Class <- gsub("Elasmobranchii", "Fish",clpi_data$Class)
clpi_data$Class <- gsub("Dipneusti", "Fish",clpi_data$Class)
clpi_data$Class <- gsub("Petromyzonti", "Fish",clpi_data$Class)
clpi_data$Class <- gsub("Holocephali", "Fish",clpi_data$Class)

# filter out fish only
fish_CLPI <- clpi_data %>% 
  filter(Class == "Fish")

# what species do we have abundance data for?
CLPI_fish_species <- unique(fish_CLPI$Binomial)

#replace '_' with ' ' to match fishbase format
CLPI_fish_species <- gsub("_", " ", CLPI_fish_species)


```
 # Trophic Trait Data 
```{r}
#Pulling trophic level traits for all Canadian fish 
CLPI_fish_trophic_data <- ecology(species_list = CLPI_fish_species,
                        fields = c("Species", "Herbivory2", "DietTroph", "DietSeTroph", 
                                   "FoodTroph", "FoodSeTroph", "DietRemark"))

## filter to species that have trophic information in at least one column 
CLPI_fish_trophic_data <- CLPI_fish_trophic_data %>% 
  filter_at(vars(Herbivory2, DietTroph, FoodTroph), any_vars(!is.na(.)))

skim(CLPI_fish_trophic_data)

hist(CLPI_fish_trophic_data$DietTroph)
```

#CLPI fish species - length data 
```{r}
CLPI_fish_size_LW <- length_weight(CLPI_fish_species) %>% 
  select(Species, LengthMin, LengthMax, Type, Sex) %>% 
  filter_at(vars(LengthMin, LengthMax), any_vars(!is.na(.))) %>% ## filter to species that have size info in at least one column 
  rename(LengthMin_LW = LengthMin, LengthMax_LW = LengthMax, Type_LW = Type, Sex_LW = Sex)

skim(CLPI_fish_size_LW)

CLPI_fish_popchar <- popchar(CLPI_fish_species) %>% 
  select(Species, PopCharRefNo, Sex, Lmax, Type, tmax) %>% 
  rename(Lmax_popchar = Lmax, Type_popchar = Type, Sex_popchar = Sex, Life_Span_popchar = tmax)

```


# CLPI fish species - life span data (from the estimate() package)
```{r}
clpi_fish_age <- estimate(CLPI_fish_species) %>% 
  select(Species, AgeMin, AgeMax) %>% 
  filter(!is.na(AgeMax))

clpi_fish_age[which(clpi_fish_age$AgeMax == max(clpi_fish_age$AgeMax)), ]


```


# Distribution of the CLPI fish traits 
```{r,fig.show='hold', message=FALSE, warning = FALSE, fig.height = 4, fig.width=9}
A <- ggplot(CLPI_fish_size_LW) +
  geom_histogram(aes(x = (LengthMax_LW)),
                 binwidth = 10, col = "white",
                 fill = "navyblue", alpha = .6) +
  labs(x = "Maximum Length (cm)")

B <- ggplot(CLPI_fish_trophic_data) +
  geom_histogram(aes(x = DietTroph),
                 binwidth = 0.5, col = "white",
                 fill = "firebrick", alpha = .6) +
  labs(x = "Trophic Level")

C <- ggplot(clpi_fish_age) +
  geom_histogram(aes(x = AgeMax),
                 bins = 25, col = "white",
                 fill = "darkgreen", alpha = .6) +
  labs(x = "Max Age")
A
B
C

```
# Comparing Trait Distributions 
```{r}
# make palette
colors <- c("Canadian Wild Species" = "navyblue",
            "C-LPI" = "firebrick")

data.frame(
  "Scale" = c("Canada", "Canada (LPI)"),
  "Species list" = c("Wild Species Report", "Canadian_LPI"),
  "Number of mammal species" = c(length(canadian_fish_sp), 
                                 length(CLPI_fish_species))
  ) %>% 
  kableExtra::kable(col.names = c("Scale", "Species list source", "Number of Mammal Species")) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")

ggplot() +
  geom_density(data = all_canadian_fish_trophic_data,
               aes(x = DietTroph, fill = "Canadian Wild Species"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = CLPI_fish_trophic_data,
               aes(x = DietTroph, fill = "C-LPI"),
               lwd = 0.2, alpha = .5) +
  scale_fill_manual(values = colors) +
  labs(title = "trophic level ", x = "trophic level", y = "Density",
       fill = "Dataset")

ggplot() +
  geom_density(data = all_canadian_fish_age,
               aes(x = AgeMax, fill = "Canadian Wild Species"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = clpi_fish_age,
               aes(x = AgeMax, fill = "C-LPI"),
               lwd = 0.2, alpha = .5) +
  scale_fill_manual(values = colors) +
  labs(title = "Maximum Age", x = "max age", y = "Density",
       fill = "Dataset")

all_canadian_fish_size_LW <- all_canadian_fish_size_LW %>% 
  filter(Type_LW == "TL")
CLPI_fish_size_LW <- CLPI_fish_size_LW %>% 
  filter(Type_LW == "TL")

ggplot() +
  geom_density(data = all_canadian_fish_size_LW,
               aes(x = LengthMax_LW, fill = "Canadian Wild Species"),
               lwd = 0.2, alpha = .5) +
  geom_density(data = CLPI_fish_size_LW,
               aes(x = LengthMax_LW, fill = "C-LPI"),
               lwd = 0.2, alpha = .5) +
  scale_fill_manual(values = colors) +
  labs(title = "Maximum Total Length", x = "max total length", y = "Density",
       fill = "Dataset")




```

