---
title: "Relationships between population change (lambdas) and traits"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: paper
---
```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

library(here)
library(dplyr)
library(lme4)
library(performance)
library(sjPlot)
library(ggplot2)
library(visdat)
library(patchwork)

# set a ggplot theme for all plots
theme_set(ggpubr::theme_pubr())

model_data = read.csv(here("scripts/CIEE_model_data.csv"), as.is = T)
model_data = model_data %>%
  filter(Replicate == 0)

trait_data = data.table::fread(here("data-clean/traits-all.csv"))

# join the datasets
df_join = left_join(model_data, trait_data, by = "Binomial")
# convert trophic level to factor
df_join$TrophicLevel <- factor(df_join$TrophicLevel)
```

# Modelling change as average lambdas

```{r, echo = FALSE}
# omit rows with NAs to ensure all models have the same data
df <- na.omit(df_join)

# plot histograms of the response variables
A <- ggplot(df) +
  geom_density(aes(x = avlambda, fill = Class), lwd = .2) +
  facet_wrap(~Class, scales = "free", ncol = 1) +
  labs(x = "Average population change (lambda)", 
       y = "Density") + theme(legend.position = "none")
B <- ggplot(df) +
  geom_density(aes(x = sumlambda, fill = Class), lwd = .2) +
  facet_wrap(~Class, scales = "free", ncol = 1) +
  labs(x = "Average population change (lambda)", y = "") + theme(legend.position = "none")
A + B + patchwork::plot_annotation(tag_levels = "a", )
ggsave(here("figures/distribution_avlambda_density.png"), width = 6.62, height = 7.5)

ggplot(df) +
  geom_boxplot(aes(y = avlambda, x = Class, fill = Class, col = Class), alpha = .5) +
  labs(y = "Average population change (lambda)", y = "")
ggsave(here("figures/distribution_avlambda_boxplot.png"), width = 6.77, height = 4.07)
```


## All taxa in Canada

These are models to evaluate the relationship between average population change and body size, trophic level, and lifespan traits, while accounting for differences in System and Class, and including a random effect by Binomial because the trait data was collected at the species level.

```{r}
# null model
m0 = lmer(avlambda ~ (1|Binomial), data = df)
m1 = lmer(avlambda ~ Class + (1|Binomial),  data = df)
m2 = lmer(avlambda ~ Class + System + (1|Binomial), data = df)
m3 = lmer(avlambda ~ System + (1|Binomial), data = df)

# adding traits

# body size only
# class should be in each model because body size is normalised within classes
m4 = lmer(avlambda ~ log10(BodySize) + Class + (1|Binomial),  data = df)
m5 = lmer(avlambda ~ log10(BodySize) + Class + System + (1|Binomial),  data = df)
m6 = lmer(avlambda ~ Class*log10(BodySize) + (1|Binomial),  data = df)
m7 = lmer(avlambda ~ System*log10(BodySize) + (1|Binomial),  data = df)

# trophic level only
m8 = lmer(avlambda ~ TrophicLevel + (1|Binomial),  data = df)
m9 = lmer(avlambda ~ TrophicLevel + Class + (1|Binomial),  data = df)
m10 = lmer(avlambda ~ TrophicLevel + System + (1|Binomial),  data = df)
m11 = lmer(avlambda ~ Class*TrophicLevel + (1|Binomial),  data = df)
m12 = lmer(avlambda ~ System*TrophicLevel + (1|Binomial),  data = df)

# lifespan only
m13 = lmer(avlambda ~ LifeSpan + (1|Binomial),  data = df)
m14 = lmer(avlambda ~ LifeSpan + Class + (1|Binomial),  data = df)
m15 = lmer(avlambda ~ LifeSpan + System + (1|Binomial),  data = df)
m16 = lmer(avlambda ~ Class*LifeSpan + (1|Binomial),  data = df)
m17 = lmer(avlambda ~ System*LifeSpan + (1|Binomial),  data = df)

# all three traits
m18 = lmer(avlambda ~ log10(BodySize) + TrophicLevel + Class + (1|Binomial),  data = df)
m19 = lmer(avlambda ~ log10(BodySize) + LifeSpan + Class + (1|Binomial),  data = df)
m20 = lmer(avlambda ~ TrophicLevel + LifeSpan + (1|Binomial),  data = df)
m21 = lmer(avlambda ~ log10(BodySize) + TrophicLevel + LifeSpan + Class + (1|Binomial),  data = df)
```

-----

### Model comparison

If we compare these models, their performance looks like...

```{r}
compare_performance(m0, m1, m2, m3, # no traits
                    m4, m5, m6, m7, # body size
                    m8, m9, m10, m11, m12, # trophic level
                    m13, m14, m15, m16, m17, # lifespan
                    m18, m19, m20, m21, # variations of 2 or 3 traits
                    rank = TRUE) %>% head() %>%
  kableExtra::kable() %>% kableExtra::kable_styling("striped")
```

-----

### "Best" model(s) evaluation

Let's look at the top models!

#### Model 12 : `avlambda ~ System*TrophicLevel + (1|Binomial)`

```{r, fig.height=9}
#check_model(m12) this is giving an error now?
```

```{r, results='hide', fig.show='hold', out.width='50%'}
#plot_model(m12, type="pred")
```

-----

## Mammals in Canada

These are models to evaluate the relationship between average population change and body size, trophic level, and lifespan traits, while accounting for differences in System, and including a random effect by Binomial because the trait data was collected at the species level. The Class variable is not included in these models, because the data is subset within the Mammals class.

```{r}
df_taxon = filter(df_join, Class == "Mammals") %>% na.omit()
```


```{r}
# null model
m0 = lmer(avlambda ~ (1|Binomial), data = df_taxon)
m1 = lmer(avlambda ~ System + (1|Binomial), data = df_taxon)

# adding traits

# body size only
# class should be in each model because body size is normalised within classes
m2 = lmer(avlambda ~ log10(BodySize) + (1|Binomial),  data = df_taxon)
m3 = lmer(avlambda ~ log10(BodySize) + System + (1|Binomial),  data = df_taxon)
m4 = lmer(avlambda ~ System*log10(BodySize) + (1|Binomial),  data = df_taxon)

# trophic level only
m5 = lmer(avlambda ~ TrophicLevel + (1|Binomial),  data = df_taxon)
m6 = lmer(avlambda ~ TrophicLevel + System + (1|Binomial),  data = df_taxon)
m7 = lmer(avlambda ~ System*TrophicLevel + (1|Binomial),  data = df_taxon)

# lifespan only
m8 = lmer(avlambda ~ LifeSpan + (1|Binomial),  data = df_taxon)
m9 = lmer(avlambda ~ LifeSpan + System + (1|Binomial),  data = df_taxon)
m10 = lmer(avlambda ~ System*LifeSpan + (1|Binomial),  data = df_taxon)

# all three traits
m11 = lmer(avlambda ~ log10(BodySize) + TrophicLevel + (1|Binomial),  data = df_taxon)
m12 = lmer(avlambda ~ log10(BodySize) + LifeSpan + (1|Binomial),  data = df_taxon)
m13 = lmer(avlambda ~ TrophicLevel + LifeSpan + (1|Binomial),  data = df_taxon)
m14 = lmer(avlambda ~ log10(BodySize) + TrophicLevel + LifeSpan + (1|Binomial),  data = df_taxon)
```

-------

### Model comparison

```{r}
compare_performance(m0, m1, # no traits
                    m2, m3, m4, # body size
                    m5, m6, m7, # trophic
                    m8, m9, m10, # lifespan
                    m11, m12, m13, m14, # combos of traits
                    rank = TRUE) %>% head() %>%
  kableExtra::kable() %>% kableExtra::kable_styling("striped")
```

-------

### "Best" model(s) evaluation

The two best models are two of the null options (m0 and m1). However, let's take a look at the best model with traits included:

#### Model 5:  `avlambda ~ TrophicLevel + (1|Binomial)`

```{r, fig.height=9}
#check_model(m5)
```

```{r, results='hide', fig.show='hold', out.width='50%'}
plot_model(m5, type="pred")
```

```{r, fig.height = 9, results='hide', fig.show='hold', out.width='50%'}
plot_model(m5, type="re")
```


#### Model 13: `avlambda ~ TrophicLevel + LifeSpan + (1|Binomial)`

```{r, fig.height=9}
#check_model(m13)
```

```{r, results='hide', fig.show='hold', out.width='50%'}
plot_model(m13, type="pred")
```

```{r, fig.height = 9, results='hide', fig.show='hold', out.width='50%'}
plot_model(m13, type="re")
```


## Birds in Canada

These are models to evaluate the relationship between average population change and body size, trophic level, and lifespan traits, while accounting for differences in System, and including a random effect by Binomial because the trait data was collected at the species level. The Class variable is not included in these models, because the data is subset within the Birds class.

```{r}
bird_df <- readRDS(here("data-clean", "LPI_birds_traits.rds"))
```

```{r}
# check missing values 
vis_miss(bird_df)
```

```{r}
bird_tidy <- model_data %>%
  left_join(bird_df, by = "Binomial") %>%
  filter(Class == "Birds") %>%
  rename(
    body_size = mean_adult_body_mass_g,
    max_lifespan = mean_max_longevity_y, 
    lifespan = mean_longevity_y, 
  ) %>%
  mutate(diet = factor(diet))

bird_traits <- c(
  "hwi", 
  "range_size", 
  "diet", 
  "body_size",
  "max_lifespan",
  "lifespan"
  ) 

# ensure complete cases for all traits 
# helps with comparing AIC values 
bird_tidy <- bird_tidy[complete.cases(bird_tidy[, bird_traits]), ]
```

```{r}
# check missing values 
vis_miss(bird_tidy)
```

```{r}
# null model
m0_b = lmer(avlambda ~ (1|Binomial), data = bird_tidy)
m1_b = lmer(avlambda ~ System + (1|Binomial), data = bird_tidy)

# adding traits

# body size only
# class should be in each model because body size is normalised within classes
m2_b = lmer(avlambda ~ log10(body_size) + (1|Binomial),  data = bird_tidy)
m3_b = lmer(avlambda ~ log10(body_size) + System + (1|Binomial),  data = bird_tidy)
m4_b = lmer(avlambda ~ System*log10(body_size) + (1|Binomial),  data = bird_tidy)

# trophic level only
m5_b = lmer(avlambda ~ diet + (1|Binomial),  data = bird_tidy)
m6_b = lmer(avlambda ~ diet + System + (1|Binomial),  data = bird_tidy)
m7_b = lmer(avlambda ~ System*diet + (1|Binomial),  data = bird_tidy)

# lifespan only
m8_b = lmer(avlambda ~ lifespan + (1|Binomial),  data = bird_tidy)
m9_b = lmer(avlambda ~ lifespan + System + (1|Binomial),  data = bird_tidy)
m10_b = lmer(avlambda ~ System*lifespan + (1|Binomial),  data = bird_tidy)

# all three traits
m11_b = lmer(avlambda ~ log10(body_size) + diet + (1|Binomial),  data = bird_tidy)
m12_b = lmer(avlambda ~ log10(body_size) + lifespan + (1|Binomial),  data = bird_tidy)
m13_b = lmer(avlambda ~ diet + lifespan + (1|Binomial),  data = bird_tidy)
m14_b = lmer(avlambda ~ log10(body_size) + diet + lifespan + (1|Binomial),  data = bird_tidy)

# hwi
m15_b <- lmer(avlambda ~ hwi + (1|Binomial), data = bird_tidy)

# combination of traits
m16_b <- lmer(avlambda ~ scale(range_size)*scale(body_size) + (1|Binomial), data = bird_tidy)
```


```{r}
compare_performance(m0_b, m1_b, m2_b, m3_b, # no traits
                    m4_b, m5_b, m6_b, m7_b, # body size
                    m8_b, m9_b, m10_b, m11_b, m12_b, # trophic level
                    m13_b, m14_b, m15_b, m16_b, # variations of 2 or 3 traits
                    rank = TRUE) %>%
  kableExtra::kable() %>% kableExtra::kable_styling("striped")
```

-------

### "Best" model(s) evaluation

Let's take a look at the best model with traits included:


#### Model 7:  `avlambda ~ System* diet + (1|Binomial)`

```{r, fig.height=9}
check_model(m7_b)
```

```{r, results='hide', fig.show='hold', out.width='50%'}
plot_model(m7_b, type="pred")
```

### Model 15:  `avlambda ~ hwi + (1|Binomial)`

```{r, fig.height=9}
check_model(m15_b)
```

```{r, results='hide', fig.show='hold', out.width='50%'}
plot_model(m15_b, type="pred")
```

------

# Modelling change as summed lambdas

```{r, echo = FALSE}
# plot histograms of the response variables
ggplot(df) +
  geom_density(aes(x = sumlambda, fill = Class), lwd = .2) +
  facet_wrap(~Class, scales = "free")
ggsave(here("figures/distribution_sumlambda_density.png"), width = 11.2, height = 4.06)

ggplot(df) +
  geom_boxplot(aes(y = sumlambda, x = Class, fill = Class, col = Class), alpha = .5)
ggsave(here("figures/distribution_sumlambda_boxplot.png"), width = 6.77, height = 4.07)
```


## All taxa in Canada

These are models to evaluate the relationship between summed population change and body size, trophic level, and lifespan traits, while accounting for differences in System and Class, and including a random effect by Binomial because the trait data was collected at the species level. Each model includes time series length to account for differences in summed lambdas in shorter vs. longer time series.

```{r}
# null model
m0 = lmer(sumlambda ~ (1|Binomial) + tslength, data = df)
m1 = lmer(sumlambda ~ Class + (1|Binomial) + tslength,  data = df)
m2 = lmer(sumlambda ~ Class + System + (1|Binomial) + tslength, data = df)
m3 = lmer(sumlambda ~ System + (1|Binomial) + tslength, data = df)

# adding traits

# body size only
# class should be in each model because body size is normalised within classes
m4 = lmer(sumlambda ~ log10(BodySize) + Class + (1|Binomial) + tslength,  data = df)
m5 = lmer(sumlambda ~ log10(BodySize) + Class + System + (1|Binomial) + tslength,  data = df)
m6 = lmer(sumlambda ~ Class*log10(BodySize) + (1|Binomial) + tslength,  data = df)
m7 = lmer(sumlambda ~ System*log10(BodySize) + (1|Binomial) + tslength,  data = df)

# trophic level only
m8 = lmer(sumlambda ~ TrophicLevel + (1|Binomial) + tslength,  data = df)
m9 = lmer(sumlambda ~ TrophicLevel + Class + (1|Binomial) + tslength,  data = df)
m10 = lmer(sumlambda ~ TrophicLevel + System + (1|Binomial) + tslength,  data = df)
m11 = lmer(sumlambda ~ Class*TrophicLevel + (1|Binomial) + tslength,  data = df)
m12 = lmer(sumlambda ~ System*TrophicLevel + (1|Binomial) + tslength,  data = df)

# lifespan only
m13 = lmer(sumlambda ~ LifeSpan + (1|Binomial) + tslength,  data = df)
m14 = lmer(sumlambda ~ LifeSpan + Class + (1|Binomial) + tslength,  data = df)
m15 = lmer(sumlambda ~ LifeSpan + System + (1|Binomial) + tslength,  data = df)
m16 = lmer(sumlambda ~ Class*LifeSpan + (1|Binomial) + tslength,  data = df)
m17 = lmer(sumlambda ~ System*LifeSpan + (1|Binomial) + tslength,  data = df)

# all three traits
m18 = lmer(sumlambda ~ log10(BodySize) + TrophicLevel + Class + (1|Binomial) + tslength,  data = df)
m19 = lmer(sumlambda ~ log10(BodySize) + LifeSpan + Class + (1|Binomial) + tslength,  data = df)
m20 = lmer(sumlambda ~ TrophicLevel + LifeSpan + (1|Binomial) + tslength,  data = df)
m21 = lmer(sumlambda ~ log10(BodySize) + TrophicLevel + LifeSpan + Class + (1|Binomial) + tslength,  data = df)
```

-------

### Model comparison

```{r}
compare_performance(m0, m1, m2, m3, # no traits
                    m4, m5, m6, m7, # body size
                    m8, m9, m10, m11, m12, # trophic level
                    m13, m14, m15, m16, m17, # lifespan
                    m18, m19, m20, m21, # variations of 2 or 3 traits
                    rank = TRUE) %>% head() %>%
  kableExtra::kable() %>% kableExtra::kable_styling("striped")
```

---------

### "Best" model(s) evaluation

#### Model 6: `sumlambda ~ Class*log10(BodySize) + (1|Binomial) + tslength`

```{r, fig.height=9}
#check_model(m6)
```

```{r, fig.show='hold', out.width='50%', results='hide'}
plot_model(m6, type="pred")
```


#### Model 16: `sumlambda ~ Class*LifeSpan + (1|Binomial) + tslength`

```{r, fig.height=9}
#check_model(m16)
```

```{r, fig.show='hold', out.width='50%', results='hide'}
plot_model(m16, type="pred")
```


## Mammals in Canada

These are models to evaluate the relationship between summed population change and body size, trophic level, and lifespan traits, while accounting for differences in System and Class, and including a random effect by Binomial because the trait data was collected at the species level. Each model includes time series length to account for differences in summed lambdas in shorter vs. longer time series.

```{r}
# null model
m0 = lmer(sumlambda ~ (1|Binomial), data = df_taxon)
m1 = lmer(sumlambda ~ System + (1|Binomial), data = df_taxon)

# adding traits

# body size only
# class should be in each model because body size is normalised within classes
m2 = lmer(sumlambda ~ log10(BodySize) + (1|Binomial) + tslength,  data = df_taxon)
m3 = lmer(sumlambda ~ log10(BodySize) + System + (1|Binomial) + tslength,  data = df_taxon)
m4 = lmer(sumlambda ~ System*log10(BodySize) + (1|Binomial) + tslength,  data = df_taxon)

# trophic level only
m5 = lmer(sumlambda ~ TrophicLevel + (1|Binomial) + tslength,  data = df_taxon)
m6 = lmer(sumlambda ~ TrophicLevel + System + (1|Binomial) + tslength,  data = df_taxon)
m7 = lmer(sumlambda ~ System*TrophicLevel + (1|Binomial) + tslength,  data = df_taxon)

# lifespan only
m8 = lmer(sumlambda ~ LifeSpan + (1|Binomial) + tslength,  data = df_taxon)
m9 = lmer(sumlambda ~ LifeSpan + System + (1|Binomial) + tslength,  data = df_taxon)
m10 = lmer(sumlambda ~ System*LifeSpan + (1|Binomial) + tslength,  data = df_taxon)

# all three traits
m11 = lmer(sumlambda ~ log10(BodySize) + TrophicLevel + (1|Binomial) + tslength,  data = df_taxon)
m12 = lmer(sumlambda ~ log10(BodySize) + LifeSpan + (1|Binomial) + tslength,  data = df_taxon)
m13 = lmer(sumlambda ~ TrophicLevel + LifeSpan + (1|Binomial) + tslength,  data = df_taxon)
m14 = lmer(sumlambda ~ log10(BodySize) + TrophicLevel + LifeSpan + (1|Binomial) + tslength,  data = df_taxon)
```

-------

### Model comparison

```{r}
compare_performance(m0, m1, # no traits
                    m2, m3, m4, # body size
                    m5, m6, m7, # trophic
                    m8, m9, m10, # lifespan
                    m11, m12, m13, m14, # combos of traits
                    rank = TRUE) %>% head() %>%
  kableExtra::kable() %>% kableExtra::kable_styling("striped")
```

---------

### "Best" model(s) evaluation

#### Model 5: `sumlambda ~ TrophicLevel + (1|Binomial) + tslength`

```{r, fig.height=9}
#check_model(m5)
```

```{r, fig.show='hold', out.width='50%', results='hide'}
plot_model(m5, type="pred")
```

## Birds in Canada

These are models to evaluate the relationship between summed population change 
and body size, trophic level, and lifespan traits, while accounting for 
differences in System, and including a random effect by Binomial because the 
trait data was collected at the species level. The Class variable is not 
included in these models, because the data is subset within the Birds class. 

Each model includes time series length to account for differences in summed 
lambdas in shorter vs. longer time series. 

```{r}
# null model
m0_sb = lmer(sumlambda ~ (1|Binomial), data = bird_tidy)
m1_sb = lmer(sumlambda ~ System + (1|Binomial), data = bird_tidy)

# adding traits

# body size only
# class should be in each model because body size is normalised within classes
m2_sb = lmer(sumlambda ~ log10(body_size) + (1|Binomial) + tslength,  data = bird_tidy)
m3_sb = lmer(sumlambda ~ log10(body_size) + System + (1|Binomial) + tslength,  data = bird_tidy)
m4_sb = lmer(sumlambda ~ System + log10(body_size) + (1|Binomial) + tslength,  data = bird_tidy)

# trophic level only
m5_sb = lmer(sumlambda ~ diet + (1|Binomial) + tslength,  data = bird_tidy)
m6_sb = lmer(sumlambda ~ diet + System + (1|Binomial) + tslength,  data = bird_tidy)
m7_sb = lmer(sumlambda ~ System*diet + (1|Binomial) + tslength,  data = bird_tidy)

# lifespan only
m8_sb = lmer(sumlambda ~ lifespan + (1|Binomial) + tslength,  data = bird_tidy)
m9_sb = lmer(sumlambda ~ lifespan + System + (1|Binomial) + tslength,  data = bird_tidy)
m10_sb = lmer(sumlambda ~ System + lifespan + (1|Binomial) + tslength,  data = bird_tidy)

# all three traits
m11_sb = lmer(sumlambda ~ log10(body_size) + diet + (1|Binomial) + tslength, data = bird_tidy)
m12_sb = lmer(sumlambda ~ log10(body_size) + lifespan + (1|Binomial) + tslength, data = bird_tidy)
m13_sb = lmer(sumlambda ~ diet + lifespan + (1|Binomial) + tslength, data = bird_tidy)
m14_sb = lmer(sumlambda ~ log10(body_size) + diet + lifespan + (1|Binomial) + tslength,  data = bird_tidy)

# hwi only
m15_sb <- lmer(sumlambda ~ hwi + (1|Binomial), data = bird_tidy)

# combination of two traits
m16_sb <- lmer(sumlambda ~ scale(range_size)*scale(body_size) + (1|Binomial), data = bird_tidy)
```


-------

### Model comparison

```{r}
compare_performance(m0_sb, m1_sb, # no traits
                    m2_sb, m3_sb, m4_sb, # body size
                    m5_sb, m6_sb, m7_sb, # trophic
                    m8_sb, m9_sb, m10_sb, # lifespan
                    m11_sb, m12_sb, m13_sb, m14_sb, # combos of traits
                    m15_sb, m16_sb, 
                    rank = TRUE) %>% head() %>%
  kableExtra::kable() %>% kableExtra::kable_styling("striped")
```

#### Model 5: `sumlambda ~ diet + (1|Binomial) + tslength`

```{r, fig.height=9}
check_model(m5_sb)
```

```{r, fig.show='hold', out.width='50%', results='hide'}
plot_model(m5_sb, type="pred")
```

```{r, fig.show='hold', out.width='50%', results='hide'}
plot_model(m5_sb, type="re")
```
#### Model 6: `sumlambda ~ diet + System + (1|Binomial) + tslength`

```{r, fig.height=9}
check_model(m6_sb)
```

```{r, fig.show='hold', out.width='50%', results='hide'}
plot_model(m6_sb, type="pred")
```