---
title: "Some attempts at modelling population response"
author: "Robin Freeman, Jessica Currie"
date: "1/09/2021"
output: 
  html_document:
    theme: united
    highlight: tango
---

## Some models

```{r setup, echo=FALSE}
library(ggpubr)
# set ggplot themes 
theme_set(theme_pubclean())
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE) 

library(tidyverse)
library(data.table)
library(stringi)

# For models
library(lme4)
library(sjPlot)
library(performance)
```

# Load Canadian LPI data and population lambdas

This first section loads the LPI and lambda data and makes a model data frame. I've switched off evaluation for now, but it's here if you need to recreate it

```{r load_lpi, eval=FALSE}

# Load selected columns from LPI data
lpi_data <- fread("data-raw/CIEE_LPI_dataset.csv", na.strings = "NA", 
                  select = c("Latitude", "Replicate", "Longitude", "ID", "System", "Class", "Binomial", "Common_name"))

##
# Simplify Class names (various Fish classes to 'Fishes', Aves/Mammalia etc to Birds/Mammals)
##
lpi_data$Taxa <- lpi_data$Class 
lpi_data$Taxa <- gsub("Actinopteri", "Fishes",lpi_data$Taxa)
lpi_data$Taxa <- gsub("Coelacanthi", "Fishes",lpi_data$Taxa)
lpi_data$Taxa <- gsub("Elasmobranchii", "Fishes",lpi_data$Taxa)
lpi_data$Taxa <- gsub("Dipneusti", "Fishes",lpi_data$Taxa)
lpi_data$Taxa <- gsub("Petromyzonti", "Fishes",lpi_data$Taxa)
lpi_data$Taxa <- gsub("Holocephali", "Fishes",lpi_data$Taxa)

lpi_data$Taxa <- gsub("Myxini", "Fishes",lpi_data$Taxa)
lpi_data$Taxa <- gsub("Aves", "Birds",lpi_data$Taxa)
lpi_data$Taxa <- gsub("Mammalia", "Mammals",lpi_data$Taxa)
lpi_data$Taxa <- gsub("Amphibia", "Amphibians",lpi_data$Taxa)
lpi_data$Taxa <- gsub("Reptilia", "Reptiles",lpi_data$Taxa)

##
# Remove replicates, excluded and "excluded 2020"
##
#lpi_data <- subset(lpi_data, Confidential == 0)

lpi_data$Latitude = as.numeric(lpi_data$Latitude)
lpi_data$Longitude = as.numeric(lpi_data$Longitude)

# Make the ID column numeric
lpi_data$ID = as.numeric(lpi_data$ID)

# Load lambda data (**you can find this file in the example_code/canadian_poplevel/ folder **)
lpi_pop_data = fread("/Users/sandraemry/Documents/example_code-main/canadian_poplevel/all_canada_pops_poplevel.csv_pops_lambda.csv", na.strings = "NA")
split_index = stri_locate_last(lpi_pop_data$V1, fixed = "_")[, 1] + 1
lpi_pop_data$ID = as.numeric(stri_sub(lpi_pop_data$V1, split_index, nchar(lpi_pop_data$V1)))

# Calculate average and total change in annual columns
lpi_pop_data$`1950` = NA # Set 1950 to NA
lpi_pop_data$sumlambda = rowSums(dplyr:::select(lpi_pop_data, starts_with(c("19", "20"))), na.rm = T)
lpi_pop_data$avlambda = rowMeans(dplyr:::select(lpi_pop_data, starts_with(c("19", "20"))), na.rm = T)
lpi_pop_data$tslength = rowSums(!is.na(dplyr:::select(lpi_pop_data, starts_with(c("19", "20")))), na.rm = T)
lpi_pop_data$ndecline = rowSums((dplyr:::select(lpi_pop_data, starts_with(c("19", "20"))) < 0), na.rm = T)


# Merge back onto large LPI dataframe
combined_data = merge(lpi_data, lpi_pop_data, by=c("ID"), all.x = TRUE)

# Set labels for increasing/declining
combined_data$Trend = "Stable"
combined_data$Trend[combined_data$sumlambda > 0.02227639] = "Increase" # Overall change of >5%
combined_data$Trend[combined_data$sumlambda < -0.02227639] = "Decline" # Overall decline of <5%

# Make Trend Taxa columns factors 
combined_data$Trend = factor(combined_data$Trend, levels = c("Decline", "Stable", "Increase"))
combined_data$Taxa = factor(combined_data$Taxa, levels = c("Fishes", "Amphibians", "Reptiles", "Birds", "Mammals"))

# Combine classes to make Herps
combined_data$Taxa = forcats::fct_collapse(combined_data$Taxa, "Herps" = c("Reptiles","Amphibians"))

# Set factor order 
combined_data$Trend = factor(combined_data$Trend, levels = c("Decline", "Stable", "Increase"))
combined_data$Taxa = factor(combined_data$Taxa, levels = c("Fishes", "Herps", "Birds", "Mammals"))

model_data = combined_data %>%
  dplyr::select(ID, Latitude, Longitude, Replicate, System, Class, Binomial, Common_name, Freq, sumlambda, avlambda,Trend, tslength, ndecline)

# Make a location factor in case we need it
model_data$location = paste(model_data$Latitude, model_data$Longitude, sep = ",")

# *** Drop 6 pops with NA lambdas (should check!!!) ***
model_data = model_data %>%
  drop_na()

write.csv(model_data, "CIEE_model_data.csv")
```

Try some preliminary plots

```{r model_data_clean}

model_data = read.csv("CIEE_model_data.csv", as.is = T)

model_data = model_data %>%
  filter(Replicate == 0)

ggplot(fish_model_data, aes(x=avlambda, fill=Class)) + 
  geom_density(alpha = 0.5) + 
  facet_wrap(~Class) + 
  geom_vline(xintercept = 0, color = "red") + 
  ggtitle("Distribution of average lambda (annual rate of change)")

ggplot(fish_model_data, aes(x=sumlambda, fill=Class)) + 
  geom_density(alpha = 0.5) + 
  facet_wrap(~Class) + 
  geom_vline(xintercept = 0, color = "red") + 
  ggtitle("Distribution of total lambda (overall annual rate of change)")

ggplot(fish_model_data, aes(x=Trend, fill=Class)) + 
  geom_bar(alpha = 0.5) + 
  facet_wrap(~Class) + 
  ggtitle("Overall population trend")

ggplot(fish_model_data, aes(x=ndecline/tslength, fill=Class)) + 
  geom_density(alpha = 0.5) + 
  facet_wrap(~Class) + 
  ggtitle("Proportion of declining years")

```
Subset fish data with traits 
```{r}
fish_traits <- fish_traits_subset %>% 
  select(Binomial, TrophicLevel, BodySize, LifeSpan) %>% 
  group_by(Binomial)

fish_traits <- distinct(fish_traits)

fish_model_data <- left_join(fish_traits, model_data, by = "Binomial")

fish_model_data <- fish_model_data %>% 
  mutate(System = factor(System),
         Class = factor(System),
         TrophicLevel = factor(TrophicLevel)) 
```


```{r models}

# model_data = model_data %>%
#   filter(Replicate == 0) %>%
#   filter(ID != 730 & ID != 3754 & ID != 921 & ID != 1036)

# Total lambda
marine_data <- fish_model_data %>% 
  filter(System == "Marine")

m0 <- lmer(sumlambda ~ Freq + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_class <- lmer(sumlambda ~ Freq + Class + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_troph <- lmer(sumlambda ~ Freq + TrophicLevel + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_lifespan <- lmer(sumlambda ~ Freq + LifeSpan + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_bodysize <- lmer(sumlambda ~ Freq + log(BodySize) + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_alltraits <- lmer(sumlambda ~ Freq + Class + TrophicLevel + LifeSpan + log(BodySize) + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_alltraits_int <- lmer(sumlambda ~ Freq + Class * TrophicLevel * LifeSpan * log(BodySize) + (1|Binomial), data = marine_data, weights = marine_data$Freq)


compare_performance(m0, m1_class, m1_troph, m1_lifespan, m1_bodysize, m1_alltraits, m1_alltraits_int)

check_model(m1_alltraits)

plot_model(m1_alltraits, type="pred")
#plot_model(m0, type="re")

plot_model(m1_inter)
```

```{r average_lambda}

# Average lambda
m0 <- lmer(avlambda ~ (1|Binomial), data = marine_data)
m1_class <- lmer(avlambda ~ Freq + Class + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_troph <- lmer(avlambda ~ Freq + TrophicLevel + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_lifespan <- lmer(avlambda ~ Freq + log(LifeSpan) + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_bodysize <- lmer(avlambda ~ Freq + log(BodySize) + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_alltraits <- lmer(avlambda ~ Freq + Class + TrophicLevel + log(LifeSpan) + log(BodySize) + (1|Binomial), data = marine_data, weights = marine_data$Freq)
m1_alltraits_int <- lmer(avlambda ~ Freq + Class * TrophicLevel * log(LifeSpan) * log(BodySize) + (1|Binomial), data = marine_data, weights = marine_data$Freq)

compare_performance(m0, m1_class, m1_troph, m1_lifespan, m1_bodysize, m1_alltraits, m1_alltraits_int, rank = TRUE)

compare_performance(m0, m1_class, m1_alltraits_int, rank = TRUE)

check_model(m1_alltraits_int)

check_outliers(m1_alltraits_int, method = c("cook", "pareto"))

plot_model(m1_alltraits_int, type="pred")
```

Filter out species that have less than 5 data points 
```{r}
marine_data_5 <- marine_data %>% 
  filter(Freq >= 5)
  
m0 <- lmer(avlambda ~ (1|Binomial), data = marine_data_5)
m1_class <- lmer(avlambda ~ Freq + Class + (1|Binomial), data = marine_data_5, weights = marine_data_5$Freq)
m1_troph <- lmer(avlambda ~ Freq + TrophicLevel + (1|Binomial), data = marine_data_5, weights = marine_data_5$Freq)
m1_lifespan <- lmer(avlambda ~ Freq + log(LifeSpan) + (1|Binomial), data = marine_data_5, weights = marine_data_5$Freq)
m1_bodysize <- lmer(avlambda ~ Freq + log(BodySize) + (1|Binomial), data = marine_data_5, weights = marine_data_5$Freq)
m1_alltraits <- lmer(avlambda ~ Freq + Class + TrophicLevel + log(LifeSpan) + log(BodySize) + (1|Binomial), data = marine_data_5, weights = marine_data_5$Freq)
m1_alltraits_int <- lmer(avlambda ~ Freq + Class * TrophicLevel * log(LifeSpan) * log(BodySize) + (1|Binomial), data = marine_data_5, weights = marine_data_5$Freq)


compare_performance(m0, m1_class,m1_troph, m1_lifespan, m1_bodysize, m1_alltraits, m1_alltraits_int, rank = TRUE)

plot_model(m1_class, type="pred")
```

Putting the freshwater fish back in and including system in the models 

```{r}
fish_model_data <- fish_model_data %>% 
  filter(!is.na(System)) %>% 
  filter(Freq >= 5)

fish_model_data <- fish_model_data %>% 
  filter_at(vars(LifeSpan, BodySize, TrophicLevel), all_vars(!is.na(.))) ## filter to species that 

m0 <- lmer(avlambda ~ Freq + (1|Binomial), data = fish_model_data, weights = fish_model_data$Freq)
m1_system <- lmer(avlambda ~ Freq + System + (1|Binomial), data = fish_model_data, weights = fish_model_data$Freq)
m1_class <- lmer(avlambda ~ Freq + Class + (1|Binomial), data = fish_model_data, weights = fish_model_data$Freq)
m1_troph <- lmer(avlambda ~ Freq + TrophicLevel + (1|Binomial), data = fish_model_data, weights = fish_model_data$Freq)
m1_lifespan <- lmer(avlambda ~ Freq + log(LifeSpan) + (1|Binomial), data = fish_model_data, weights = fish_model_data$Freq)
m1_bodysize <- lmer(avlambda ~ Freq + log(BodySize) + (1|Binomial), data = fish_model_data, weights = fish_model_data$Freq)
m1_alltraits <- lmer(avlambda ~ Freq + Class + TrophicLevel + log(LifeSpan) + log(BodySize) + (1|Binomial), data = fish_model_data, weights = fish_model_data$Freq)
m1_alltraits_int <- lmer(avlambda ~ Freq + Class * TrophicLevel * log(LifeSpan) * log(BodySize) + (1|Binomial), data = fish_model_data, weights = fish_model_data$Freq)

compare_performance(m0, m1_class, m1_troph, m1_lifespan, m1_bodysize, m1_alltraits, m1_alltraits_int, rank = TRUE)

## Not including the weights 
m0 <- lmer(avlambda ~ (1|Binomial), data = fish_model_data)
m1_class <- lmer(avlambda ~ Class + (1|Binomial), data = fish_model_data)
m1_system <- lmer(avlambda ~ System + (1|Binomial), data = fish_model_data)
m1_troph <- lmer(avlambda ~ TrophicLevel + (1|Binomial), data = fish_model_data)
m1_lifespan <- lmer(avlambda ~ log(LifeSpan) + (1|Binomial), data = fish_model_data)
m1_bodysize <- lmer(avlambda ~ log(BodySize) + (1|Binomial), data = fish_model_data)
m1_alltraits <- lmer(avlambda ~  Class + TrophicLevel + log(LifeSpan) + log(BodySize) + (1|Binomial), data = fish_model_data)
m1_alltraits_int <- lmer(avlambda ~  Class * TrophicLevel * log(LifeSpan) * log(BodySize) + (1|Binomial), data = fish_model_data)

compare_performance(m0, m1_class, m1_system, m1_troph, m1_lifespan, m1_bodysize, m1_alltraits, m1_alltraits_int, rank = TRUE)

```


```{r average_lambda_transformed}
# Can also transform lambda
library(LambertW)
model_data$avlambda_trans = Gaussianize(model_data$avlambda)

ggplot(model_data, aes(x = avlambda)) + geom_histogram()
ggplot(model_data, aes(x = avlambda_trans)) + geom_histogram()

# Average lambda
m0_trans = lmer(avlambda_trans ~ (1|Binomial),  data = model_data)
m1_class_trans = lmer(avlambda_trans ~ Class + (1|Binomial), data = model_data)
m1_system_trans = lmer(avlambda_trans ~ System + (1|Binomial), data = model_data)
m1_add_trans = lmer(avlambda_trans ~ System + Class + (1|Binomial),  data = model_data)
m1_inter_trans = lmer(avlambda_trans ~ System*Class + (1|Binomial),  data = model_data)

compare_performance(m0_trans, m1_class_trans, m1_system_trans, m1_add_trans, m1_inter_trans)
check_model(m1_class_trans)

plot_model(m1_class_trans, type="pred")
```

```{r glmer_models}

fish_model_data <- fish_model_data %>% 
  filter(!is.na(System)) %>% 
  filter(Freq >= 5)

fish_model_data <- fish_model_data %>% 
  filter_at(vars(LifeSpan, BodySize, TrophicLevel), all_vars(!is.na(.))) ## filter to species that have size info in at least one column 

# Likelihood of declines
glm0 = glmer(cbind(ndecline, (tslength - ndecline)) ~ (1|Binomial),  
                   family="binomial", control = glmerControl(optimizer = "bobyqa"), data = fish_model_data)

glm1_class = glmer(cbind(ndecline, (tslength - ndecline)) ~ Class + (1|Binomial),  
                   family="binomial", control = glmerControl(optimizer = "bobyqa"), data = fish_model_data)

glm1_system = glmer(cbind(ndecline, (tslength - ndecline)) ~ System + (1|Binomial),  
                   family="binomial", control = glmerControl(optimizer = "bobyqa"), data = fish_model_data)

glm1_lifespan = glmer(cbind(ndecline, (tslength - ndecline)) ~ log(LifeSpan) + (1|Binomial),  
                   family="binomial", control = glmerControl(optimizer = "bobyqa"), data = fish_model_data)

glm1_troph = glmer(cbind(ndecline, (tslength - ndecline)) ~ TrophicLevel + (1|Binomial),  
                   family="binomial", control = glmerControl(optimizer = "bobyqa"), data = fish_model_data)

glm1_bodysize = glmer(cbind(ndecline, (tslength - ndecline)) ~ log(BodySize) + (1|Binomial),  
                   family="binomial", control = glmerControl(optimizer = "bobyqa"), data = fish_model_data)

glm1_all_add <- glmer(cbind(ndecline, (tslength - ndecline)) ~ System + Class + log(LifeSpan) + TrophicLevel + log(BodySize) + (1|Binomial), family="binomial", control = glmerControl(optimizer = "bobyqa"), data = fish_model_data)
  
glm1_all_int <- glmer(cbind(ndecline, (tslength - ndecline)) ~ System * Class * log(LifeSpan) + TrophicLevel * log(BodySize) * (1|Binomial), family="binomial", control = glmerControl(optimizer = "bobyqa"), data = fish_model_data)

compare_performance(glm0, glm1_class, glm1_system, glm1_troph, glm1_lifespan, glm1_bodysize, glm1_all_add, glm1_all_int)

check_model(glm1_bodysize)

check_outliers(glm1_bodysize, method = c("cook", "pareto"))

plot_model(glm1_bodysize, type="pred")

tab_model(glm1_all_add)


```

```{r check_influential_sumlambda, eval=FALSE}

# When looking at sumlambda - these came up as strongly influential...

model_data[612, ] # ID: 730, Cygnus_buccinator, Trumpeter swan (sum: 2.255713, av: 0.04903724)
model_data[3070, ] # ID: 3754, Oncorhynchus_nerka, Sockeye salmon / Red salmon (sum: -12.51969, av: -0.2911556)
model_data[799, ] # ID: 921, Larus_fuscus, Lesser black-backed gull (sum: 3.385108, av: 0.07358931)
model_data[909, ] # ID: 1036, Colinus_virginianus, Northern bobwhite (sum: -3.663572, av: -0.08326301)

ggplot(model_data, aes(x=sumlambda, fill=Class)) + 
  geom_density(alpha = 0.5) + 
  facet_wrap(~Class) + 
  geom_vline(xintercept = 3.255713) + 
  geom_vline(xintercept = -11.51969) + 
  geom_vline(xintercept = 4.385108) + 
  geom_vline(xintercept = -2.663572) 
  

```