---
title: "Extracting IUCN Habitats"
author: "Robin Freeman, Indicators and Assessments Unit, ZSL"
date: "1/09/2021"
output: 
  html_document:
    theme: united
    highlight: tango
---

## Extracting IUCN Habitat data for Canadian species in the Living Planet Database


```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r libraries}
library(data.table)
library(ggplot2)
library(rredlist)

theme_set(theme_pubclean())
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE) 

```


```{r load_data}

# Load the Canadian species data with RL id mapping
CIEE_LPI_dataset_RL_info = fread("CIEE_LPI_dataset_RL_info.csv")

```

## Extract habitat data

For each species, we can then use the IUCN taxon ID to extract habitat information for that species. This can be slow and uses the [IUCN Red List API](https://apiv3.iucnredlist.org/) and [rredlist](https://github.com/ropensci/rredlist) package. Using this requires a Red List API key that can be requested [here](https://apiv3.iucnredlist.org/api/v3/token)

```{r get_iucn_habitat_data, eval=FALSE}

# Long and slow
habitat_data_all = list()

# Get number of species
N_species = length(CIEE_LPI_dataset_RL_info$RL_ID)

# For each species
for (i in 1:N_species) {
  # Get the red list ID
  red_list_id = as.numeric(CIEE_LPI_dataset_RL_info$RL_ID[i])
  # If it's not missing
  if (!is.na(red_list_id)) {
    # Get the binomial (just for printing here)
    species = CIEE_LPI_dataset_RL_info$RL_binomial[i]
    # Print out what we're doing
    print(sprintf("[%d of %d] Searching for: %s", i, N_species, species))
    
    # Get the habita data from Red List API using rl_habitats from rredlist
    habitat_data = rl_habitats(id = red_list_id)
    
    #print(habitat_data)
    
    # Save to list
    habitat_data_all[[i]] = habitat_data
  }
}

# Save results to RDS
saveRDS(habitat_data_all, "habitat_data_canada.Rds")
```

Load the saved list from RDS and convert into nicer tabular format, then merge back into original data frame

```{r}

habitat_data_all = readRDS("habitat_data_canada.Rds")

# Function to extract species results, add ID to data frame and return
get_results = function(x) {
  #print(x)
  if (!is.null(x) & length(x$result) > 0) {
    res = x$result
    res$id = x$id
    return(res)
  } else {
    return(NULL)
  }
}

# Apply function
habitat_data_all_results = lapply(habitat_data_all, get_results)

# Convert to data frame
habitat_data_all_df = rbindlist(habitat_data_all_results)

# Merge back into original data
habitat_data_all_df = left_join(CIEE_LPI_dataset_RL_info, habitat_data_all_df, by = c("RL_ID" = "id"))

fwrite(habitat_data_all_df, "habitat_data_all_df.csv")
```

Let's combine back to original LPI data

```{r plot_data}
habitat_data_all_df = fread("habitat_data_all_df.csv")

# Function to extract higher-level habitat code (e.g. "10" from "10.2", "10.4" etc.)
get_highlevel_code = function(x) {
  # If there's a decimal place...
  dp = str_locate(x, "\\.")[1]
  if (!is.na(dp)) {
    hl_code = str_sub(x, 1, dp - 1)
  } else {
    hl_code = x
  }
  return(hl_code)
}

# Get higher levels codes
habitat_data_all_df$hl_habitat_code = get_highlevel_code(habitat_data_all_df$code)

# Create summary data frame with Species, ID, Red List ID, System and Class
# alongside total #of unique habitats and total # of high-level habitats
species_habitat_simple = habitat_data_all_df %>%
  group_by(Binomial, RL_ID, System, Class.x) %>%
  filter(!is.null(RL_ID)) %>%
  filter(RL_ID != "NULL") %>%
  drop_na() %>%
  summarise(n_habitats = n_distinct(code),
            n_hl_habitats = n_distinct(hl_habitat_code))

# Save to file
fwrite(species_habitat_simple, "canada_species_habitat_simple.csv")

# Combine back onto original data 
canadian_lpi_data <- fread("data/CIEE_LPI_dataset.csv", na.strings = "NA")

canadian_lpi_data_with_habitat = left_join(canadian_lpi_data, 
          select(species_habitat_simple, Binomial, RL_ID, n_habitats, n_hl_habitats), 
          by="Binomial")

# And save
fwrite(canadian_lpi_data_with_habitat, "canadian_lpi_data_with_habitat.csv")
```

make some graphs (not very exciting)

```{r graphs}
ggplot(species_habitat_simple, 
       aes(x = n_habitats, fill=System)) + 
  geom_density(lwd = 0.2, alpha = .5)  + 
  facet_wrap(~System)

ggplot(species_habitat_simple, 
       aes(x = n_hl_habitats, fill=System)) + 
  geom_histogram(lwd = 0.2)  + facet_wrap(~System)


```


```{r get_iucn_threat_data, eval=FALSE}

# Long and slow
threat_data_all = list()

# Get number of species
N_species = length(CIEE_LPI_dataset_RL_info$RL_ID)

# For each species
for (i in 1:N_species) {
  # Get the red list ID
  red_list_id = as.numeric(CIEE_LPI_dataset_RL_info$RL_ID[i])
  # If it's not missing
  if (!is.na(red_list_id)) {
    # Get the binomial (just for printing here)
    species = CIEE_LPI_dataset_RL_info$RL_binomial[i]
    # Print out what we're doing
    print(sprintf("[%d of %d] Searching for: %s", i, N_species, species))
    
    # Get the habita data from Red List API using rl_habitats from rredlist
    threat_data = rl_threats(id = red_list_id)
    
    print(threat_data)
    
    # Save to list
    threat_data_all[[i]] = threat_data
  }
}

# Save results to RDS
saveRDS(threat_data_all, "threat_data_canada.Rds")
```







