---
title: "Trait distribution comparisons"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}

#read in packages and functions

library(tidyverse)
library(data.table)
library(ggpubr)
library(effectsize)
library(wesanderson)
library(traitdata)

#input:dataframe and traitname matching a column in the dataframe
#output: cohen's d
trait_stats<-function(traitname){
df.out<-data.frame(taxon=character(),KS_stat=numeric(),p_value=numeric(),Cohens_D=numeric(),a=character())
for(x in levels(as.factor(verts$Group))){
df<-subset(verts,verts$Group==x)
df<-select(df,c(traitname,"lpi"))
  colnames(df)<-c("trait","lpi")
  df.wild<-df[df$lpi=="C-wild",]
  df.lpi<-df[df$lpi=="C-LPI",]
  ks1<-ks.test(df.lpi$trait,df.wild$trait)
  cd1<-(mean(df.lpi$trait,na.rm=T)-mean(df.wild$trait,na.rm=T))/sd_pooled(trait~lpi,data=df)
  df.out<-rbind(df.out,cbind(taxon=x,KS_stat=ks1$statistic,p_value=ks1$p.value,Cohens_D=cd1,a=""))}
rownames(df.out)<-NULL
df.out$a[as.numeric(df.out$p_value)<0.05]<-"*"
df.out}

trait_stats<-function(traitname,all_sp){
  if(all_sp==F){verts<-verts_2}
  df.out<-data.frame(taxon=character(),KS_stat=numeric(),p_value=numeric(),Cohens_D=numeric(),a=character())
  for(x in c("all",levels(as.factor(verts$Group)))){
    if(x!="all"){df<-subset(verts,verts$Group==x)}
    if(x=="all"){df<-verts}
    df<-select(df,c(traitname,"lpi"))
      colnames(df)<-c("trait","lpi")
      if(all_sp==T){df.wild<-df[df$lpi=="C-wild",]}
      if(all_sp==F){df.wild<-df[df$lpi=="C-wild-only",]}
      df.lpi<-df[df$lpi=="C-LPI",]
      ks1<-ks.test(df.lpi$trait,df.wild$trait)
      cd1<-(mean(df.lpi$trait,na.rm=T)-mean(df.wild$trait,na.rm=T))/sd_pooled(trait~lpi,data=df)
      df.out<-rbind(df.out,cbind(taxon=x,KS_stat=ks1$statistic,p_value=ks1$p.value,Cohens_D=cd1,a=""))}
  rownames(df.out)<-NULL
  df.out$a[as.numeric(df.out$p_value)<0.05]<-"*"
  df.out}


trait_stats_discrete<-function(traitname,all_sp){
  if(all_sp==F){verts<-verts_2}
  df.out<-data.frame(taxon=character(),chi_sq=numeric(),p_value=numeric(),a=character())
  for(x in c("all",levels(as.factor(verts$Group)))){
    if(x!="all"){df<-subset(verts,verts$Group==x)}
    if(x=="all"){df<-verts}
    df<-select(df,c(traitname,"lpi"))
      colnames(df)<-c("trait","lpi")
      cs1<-chisq.test(x=df$trait,y=df$lpi)
      df.out<-rbind(df.out,cbind(taxon=x,chi_sq=cs1$statistic,p_value=cs1$p.value,a=""))}
  rownames(df.out)<-NULL
  df.out$a[as.numeric(df.out$p_value)<0.05]<-"*"
  df.out}


ggplot <- function(...) ggplot2::ggplot(...) + scale_color_manual(values=c("gray40","coral3")) + 
  scale_fill_manual(values=c("gray40","coral3"))

```


```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}

#set up data for Canadian + C-LPI species

herps<-fread("../data-clean/herps_canadian_sp.csv")
fish<-fread("../data-clean/fish/fishbase_traits-for-comparison.csv")
birds<-fread("../data-clean/birds_traits_globalLPI.csv")

lpi.dat<-fread("../data-raw/CIEE_LPI_dataset.csv")
wild.dat<-fread("../data-raw/WildSpecies2015Data.csv")

fish<-subset(fish,fish$collection!="All species")
fish$collection[fish$collection=="Canadian Wild Species"]<-"C-wild"
fish<-fish %>% select(c("Binomial","MaxLength_TLonly","LongevityWild",
                                "TrophCategorical","collection")) %>%
  rename(BodySize=MaxLength_TLonly,LifeSpan=LongevityWild,TrophicLevel=TrophCategorical,lpi=collection) %>%
  unique()

birds$lpi<-"Global"
birds$lpi[birds$binomial%in%wild.dat$Binomial]<-"C-wild"
birds$lpi[birds$binomial%in%lpi.dat$Binomial]<-"C-LPI"
birds<-birds %>% select(c("binomial","mean_adult_body_mass_g","mean_max_longevity_y",
                                  "diet","lpi")) %>%
  rename(Binomial=binomial,BodySize=mean_adult_body_mass_g,
         LifeSpan=mean_max_longevity_y,TrophicLevel=diet) %>%
  filter(lpi!="Global")
temp1<-birds[birds$lpi=="C-LPI",]
temp1$lpi<-"C-wild"
birds<-rbind(birds,temp1); rm(temp1)
#reassign diet categories
birds$TrophicLevel[birds$TrophicLevel%in%
                         c("fruit","nectar","plants","seeds")]<-"herbivore"
birds$TrophicLevel[birds$TrophicLevel%in%
                         c("invertebrates","vertebrates")]<-"carnivore"
birds$TrophicLevel[birds$TrophicLevel=="scav"]<-"carnivore" #turkey vulture



herps$lpi[herps$lpi=="lpi"]<-"C-LPI"
herps$lpi[herps$lpi=="not_lpi"]<-"C-wild"
herps<-herps %>% select(c("Binomial","body_mass_g","longevity_years",
                                  "diet","lpi")) %>%
  rename(BodySize=body_mass_g,LifeSpan=longevity_years,TrophicLevel=diet)
temp1<-herps[herps$lpi=="C-LPI",]
temp1$lpi<-"C-wild"
herps<-rbind(herps,temp1); rm(temp1)
herps$TrophicLevel[herps$TrophicLevel=="Herbivore"]<-"herbivore"



###MAMMALS
#following code in mammal-traits.Rmd as of 03-Sept-2021

# load pantheria dataset
data("pantheria")
# subset to canada
pantheria_canada <- pantheria %>%
  filter(scientificNameStd %in% wild.dat$`SCIENTIFIC NAME - NOM SCIENTIFIQUE`) %>%
  select(c("scientificNameStd","AdultBodyMass_g","MaxLongevity_m","TrophicLevel")) %>%
  rename(Binomial=scientificNameStd)
pantheria_canada$Binomial<-gsub(" ","_",pantheria_canada$Binomial)
     
pantheria_clpi <- readRDS("../data-clean/LPI_pantheria.rds") %>%
  filter(Class %in% c("Mammalia", "Mammals")) %>%
  # keep unique binomials only, to avoid having duplicates that falsely inflate
  # the number of observations per trait value
  distinct(Binomial, .keep_all = TRUE) %>%
  select(c("Binomial","AdultBodyMass_g","MaxLongevity_m","TrophicLevel"))
pantheria_canada$lpi<-"C-wild"
pantheria_clpi$lpi<-"C-LPI"
mammals<-rbind(pantheria_canada,pantheria_clpi) %>%
  rename(BodySize=AdultBodyMass_g,LifeSpan=MaxLongevity_m)
mammals$TrophicLevel<-gsub(1,"herbivore",mammals$TrophicLevel)
mammals$TrophicLevel<-gsub(2,"omnivore",mammals$TrophicLevel)
mammals$TrophicLevel<-gsub(3,"carnivore",mammals$TrophicLevel)


birds$Group<-"Birds"
fish$Group<-"Fish"
herps$Group<-"Herpetofauna"
mammals$Group<-"Mammals"


verts<-rbind(birds,fish,herps,mammals)
verts$BodySize.log<-log(verts$BodySize)
verts$LifeSpan.log<-log(verts$LifeSpan)


C_wild_only<-setdiff(verts$Binomial[verts$lpi=="C-wild"],verts$Binomial[verts$lpi=="C-LPI"])
verts_2<-verts[verts$Binomial%in%C_wild_only,]
verts_2<-rbind(verts_2,verts[verts$lpi=="C-LPI",])
verts_2$lpi[verts_2$lpi=="C-wild"]<-"C-wild-only"


verts$lpi<-factor(verts$lpi,levels=c("C-wild","C-LPI"))
verts_2$lpi<-factor(verts_2$lpi,levels=c("C-wild-only","C-LPI"))

rm(birds,fish,herps,mammals,pantheria,pantheria_canada,pantheria_clpi)


```

## Data checks

### Species in the LPI dataset but not the Canadian wild species list

These species (and their trait data) were added to wild species set for the trait comparisons

```{r, echo=FALSE, message=FALSE, warning=FALSE}

verts.clpi<-subset(verts,verts$lpi=="C-LPI")
verts.cwild<-subset(verts,verts$lpi=="C-wild")


##Animals in LPI but not wild database
o1<-setdiff(verts.clpi$Binomial,verts.cwild$Binomial)
verts[verts$Binomial%in%o1,c(1,5:6)]
o2<-verts[verts$Binomial%in%o1,]
o2$lpi<-"C-wild"
verts<-rbind(verts,o2)

```

### Test of normality

```{r, echo=FALSE, message=FALSE, warning=FALSE}

#used only wild species to avoid repeated values with lpi dataset

ggqqplot(subset(verts,verts$lpi=="C-wild")$BodySize.log)

shapiro.test(subset(verts,verts$lpi=="C-wild")$BodySize.log)

ggqqplot(subset(verts,verts$lpi=="C-wild")$LifeSpan.log)

shapiro.test(subset(verts,verts$lpi=="C-wild")$LifeSpan.log)

```

### Trait relationships

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width = 12}

A1<-ggplot(verts,aes(x=BodySize.log,y=LifeSpan.log,colour=lpi))+
  geom_smooth(size=2,method="lm")+
  theme_classic()+
  xlab("Body Size (log)")+ylab("Lifespan (log)")+
  theme(legend.title=element_blank())

verts$tr.levels<-paste(verts$TrophicLevel,verts$lpi,sep="_")

trophic.life.summary<-
  data.frame(means=tapply(verts$LifeSpan.log,verts$tr.levels,mean,na.rm=T),
             sds=tapply(verts$LifeSpan.log,verts$tr.levels,sd,na.rm=T))[c(1:4,7,8),]
trophic.life.summary$sd.l<-trophic.life.summary$means-trophic.life.summary$sds
trophic.life.summary$sd.u<-trophic.life.summary$means+trophic.life.summary$sds
trophic.life.summary$TrophicLevel<-t(data.frame(strsplit(rownames(trophic.life.summary),"_")))[,1]
trophic.life.summary$lpi<-t(data.frame(strsplit(rownames(trophic.life.summary),"_")))[,2]


B1<-ggplot(trophic.life.summary,aes(x=TrophicLevel,y=means,colour=lpi,group=lpi))+
  geom_point(size=4,position=position_dodge(width=0.2))+
  geom_errorbar(aes(ymin=sd.l,ymax=sd.u),width=0.2,position=position_dodge(width=0.2))+
  theme_classic()+ylab("Lifespan (log)")+xlab("Trophic Level")+
  theme(legend.title=element_blank())


trophic.body.summary<-
  data.frame(means=tapply(verts$BodySize.log,verts$tr.levels,mean,na.rm=T),
             sds=tapply(verts$BodySize.log,verts$tr.levels,sd,na.rm=T))[c(1:4,7,8),]
trophic.body.summary$sd.l<-trophic.body.summary$means-trophic.body.summary$sds
trophic.body.summary$sd.u<-trophic.body.summary$means+trophic.body.summary$sds
trophic.body.summary$TrophicLevel<-t(data.frame(strsplit(rownames(trophic.body.summary),"_")))[,1]
trophic.body.summary$lpi<-t(data.frame(strsplit(rownames(trophic.body.summary),"_")))[,2]


C1<-ggplot(trophic.body.summary,aes(x=TrophicLevel,y=means,colour=lpi,group=lpi))+
  geom_point(size=4,position=position_dodge(width=0.2))+
  geom_errorbar(aes(ymin=sd.l,ymax=sd.u),width=0.2,position=position_dodge(width=0.2))+
  theme_classic()+ylab("Body Size (log)")+xlab("Trophic Level")+
  theme(legend.title=element_blank())

ggarrange(A1,B1,C1,nrow=1,ncol=3)


```




### Data coverage for species in the LPI dataset

```{r, echo=FALSE, message=FALSE, warning=FALSE}

verts.clpi$BS.pa<-if_else(is.na(verts.clpi$BodySize.log),0,1)
verts.clpi$LS.pa<-if_else(is.na(verts.clpi$LifeSpan.log),0,1)
verts.clpi$TL.pa<-if_else(is.na(verts.clpi$TrophicLevel),0,1)


v.BS<-verts.clpi %>% group_by(Group,BS.pa) %>%  summarise(n=n()) %>%  mutate(freq=n/sum(n)) %>% 
  subset(BS.pa==1) %>% select(c("Group","n","freq"))
v.LS<-verts.clpi %>% group_by(Group,LS.pa) %>%  summarise(n=n()) %>%  mutate(freq=n/sum(n)) %>% 
  subset(LS.pa==1) %>% select(c("Group","n","freq"))
v.TL<-verts.clpi %>% group_by(Group,TL.pa) %>%  summarise(n=n()) %>%  mutate(freq=n/sum(n)) %>% 
  subset(TL.pa==1) %>% select(c("Group","n","freq"))

v.BS$Trait<-"BodySize"; v.LS$Trait<-"LifeSpan"; v.TL$Trait<-"TrophicLevel"

v.trait<-rbind(v.BS,v.LS,v.TL)


ggplot(data=v.trait,aes(x=Trait,y=freq,fill=Group))+
  geom_col(position="dodge")+theme_classic()+ylab("Frequency")+
  scale_fill_manual(values=wes_palette("Moonrise2",4))





```

### Data coverage for Canadian wild species (not LPI)

```{r, echo=FALSE, message=FALSE, warning=FALSE}

verts.cwild_only<-verts_2[verts_2$lpi=="C-wild-only",]

verts.cwild_only$BS.pa<-if_else(is.na(verts.cwild_only$BodySize.log),0,1)
verts.cwild_only$LS.pa<-if_else(is.na(verts.cwild_only$LifeSpan.log),0,1)
verts.cwild_only$TL.pa<-if_else(is.na(verts.cwild_only$TrophicLevel),0,1)


v.BS<-verts.cwild_only %>% group_by(Group,BS.pa) %>%  summarise(n=n()) %>%  mutate(freq=n/sum(n)) %>% 
  subset(BS.pa==1) %>% select(c("Group","n","freq"))
v.LS<-verts.cwild_only %>% group_by(Group,LS.pa) %>%  summarise(n=n()) %>%  mutate(freq=n/sum(n)) %>% 
  subset(LS.pa==1) %>% select(c("Group","n","freq"))
v.TL<-verts.cwild_only %>% group_by(Group,TL.pa) %>%  summarise(n=n()) %>%  mutate(freq=n/sum(n)) %>% 
  subset(TL.pa==1) %>% select(c("Group","n","freq"))

v.BS$Trait<-"BodySize"; v.LS$Trait<-"LifeSpan"; v.TL$Trait<-"TrophicLevel"

v.trait<-rbind(v.BS,v.LS,v.TL)


ggplot(data=v.trait,aes(x=Trait,y=freq,fill=Group))+
  geom_col(position="dodge")+theme_classic()+ylab("Frequency")+
  scale_fill_manual(values=wes_palette("Moonrise2",4))





```


## Trait comparisons

Canadian wild (including LPI species) compared to LPI species

### Body size

By taxonomic group

```{r, echo=FALSE, message=FALSE, warning=FALSE}



ggplot(verts,aes(x=BodySize.log,colour=lpi))+
  geom_density(size=2)+
  facet_wrap(vars(Group),nrow=2)+
  theme_classic()+
  xlab("Body Size (log)")+
  theme(legend.title=element_blank())



```

All taxa

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 5}
A2<-ggplot(verts,aes(x=BodySize.log,colour=lpi))+
  geom_density(size=2)+
  theme_classic()+
  xlab("Body Size (log)")+
  theme(legend.title=element_blank())

A2

trait_stats("BodySize.log",T)

```



Question here: I used Cohen's D to get an effect size for continuous trait differences, but I'm not sure that that's 
appropriate given that the data aren't normal. What would be a better option? I could just report the medians of each
group?

### Lifespan

By taxonomic group

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(verts,aes(x=LifeSpan.log,colour=lpi))+
  geom_density(size=2)+
  facet_wrap(vars(Group),nrow=2)+
  theme_classic()+
  xlab("Lifespan (log)")+
  theme(legend.title=element_blank())
```

All taxa

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 5}



B2<-ggplot(verts,aes(x=LifeSpan.log,colour=lpi))+
  geom_density(size=2)+
  theme_classic()+
  xlab("Lifespan (log)")+
  theme(legend.title=element_blank())
B2

trait_stats("LifeSpan.log",T)



```

### Trophic level

By taxonomic group

```{r echo=FALSE, message=FALSE, warning=FALSE}

troph.plot<-data.frame(table(verts$TrophicLevel,verts$Group,verts$lpi))
colnames(troph.plot)<-c("TrophicLevel","Group","lpi","proportion")
troph.plot$TrophicLevel<-factor(troph.plot$TrophicLevel,levels=c("herbivore","omnivore","carnivore"))

ggplot(troph.plot,aes(x=lpi,y=proportion,fill=TrophicLevel))+
  geom_col(position="fill")+
  facet_wrap(vars(Group),nrow=2)+
  scale_fill_manual(values=c("slategray1","slategray3","slategray4"))+
  theme_classic()+
  xlab("Dataset")+
  theme(legend.title=element_blank())

```

All taxa

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 5}

C2<-ggplot(troph.plot,aes(x=lpi,y=proportion,fill=TrophicLevel))+
  geom_col(position="fill")+
  scale_fill_manual(values=c("slategray1","slategray3","slategray4"))+
  xlab("Dataset")+
  theme_classic()+
  theme(legend.title=element_blank())

C2

trait_stats_discrete("TrophicLevel",T)

```


### Combined figure

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 12}



ggarrange(A2,B2,C2,nrow=1,ncol=3,labels=c("A","B","C"))



```



## Trait comparisons - Canadian LPI to not LPI

Canadian wild (NOT including LPI species) compared to LPI species

### Body size

By taxonomic group

```{r, echo=FALSE, message=FALSE, warning=FALSE}



ggplot(verts_2,aes(x=BodySize.log,colour=lpi))+
  geom_density(size=2)+
  facet_wrap(vars(Group),nrow=2)+
  theme_classic()+
  xlab("Body Size (log)")+
  theme(legend.title=element_blank())



```

All taxa

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 5}
A2<-ggplot(verts_2,aes(x=BodySize.log,colour=lpi))+
  geom_density(size=2)+
  theme_classic()+
  xlab("Body Size (log)")+
  theme(legend.title=element_blank())

A2

trait_stats("BodySize.log",F)

```

### Lifespan

By taxonomic group

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(verts_2,aes(x=LifeSpan.log,colour=lpi))+
  geom_density(size=2)+
  facet_wrap(vars(Group),nrow=2)+
  theme_classic()+
  xlab("Lifespan (log)")+
  theme(legend.title=element_blank())
```

All taxa

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 5}



B2<-ggplot(verts_2,aes(x=LifeSpan.log,colour=lpi))+
  geom_density(size=2)+
  theme_classic()+
  xlab("Lifespan (log)")+
  theme(legend.title=element_blank())
B2

trait_stats("LifeSpan.log",F)



```

### Trophic level

By taxonomic group

```{r echo=FALSE, message=FALSE, warning=FALSE}

troph.plot<-data.frame(table(verts_2$TrophicLevel,verts_2$Group,verts_2$lpi))
colnames(troph.plot)<-c("TrophicLevel","Group","lpi","proportion")
troph.plot$TrophicLevel<-factor(troph.plot$TrophicLevel,levels=c("herbivore","omnivore","carnivore"))

ggplot(troph.plot,aes(x=lpi,y=proportion,fill=TrophicLevel))+
  geom_col(position="fill")+
  facet_wrap(vars(Group),nrow=2)+
  scale_fill_manual(values=c("slategray1","slategray3","slategray4"))+
  theme_classic()+
  xlab("Dataset")+
  theme(legend.title=element_blank())

```

All taxa

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 5}

C2<-ggplot(troph.plot,aes(x=lpi,y=proportion,fill=TrophicLevel))+
  geom_col(position="fill")+
  scale_fill_manual(values=c("slategray1","slategray3","slategray4"))+
  xlab("Dataset")+
  theme_classic()+
  theme(legend.title=element_blank())

C2

trait_stats_discrete("TrophicLevel",F)

```


### Combined figure

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 12}



ggarrange(A2,B2,C2,nrow=1,ncol=3,labels=c("A","B","C"))



```





